Fractional Types

### 2019-11-18 fracGC/p.tex
paper

\paragraph*{Conservation of Information and Negative Entropy.}  
According to the conventional theory of
communication~\cite{Shannon1948}, a type with $N$ values is viewed as
an abstract system that has $N$ distinguishable states where the
amount of information contained in each state is $\log{N}$. This
entropy is a measure of information which materializes itself in
memory or bandwidth requirements when storing or transmitting elements
of this type. Thus a type with 8 elements needs 3 bits of memory for
storage or 3 bits of bandwidth for communication. The logarithmic map
implies that information contained in a composite state is the sum of
the information contained in its constituents. For example, the type
$A \times B$ where $A$ has two elements and $B$ has three elements can
be thought of a composite system consisting of two independent
unrelated subsystems.  Each state of the composite system therefore
contains $\log{(2*3)} = \log{2} + \log{3}$ bits which is the sum of
the information contained in each subsystem. If we could imagine a
\emph{fractional type} $\frac{1}{A}$, this type would introduce
negative entropy. For example, a type with ``cardinality''
$\frac{1}{8}$ has entropy $\log{\frac{1}{8}} = -3$. It is natural to
interpret this negative entropy just like we interpret ``negative
money,'' as a resource (space or bandwidth) to be repaid (reclaimed)
by some other part of the system. Indeed, we will introduce such
fractional types and use them to represent ``garbage collection
processes'' that reclaim temporary storage. Just like in the case of
``negative money'' it is critical that these fractional types be
linearly used, i.e., they cannot be duplicated or erased: this
property is however automatically enforced in a reversible
computational model!

\paragraph*{Outline.} The paper solves the ancilla problem in
reversible and quantum computation using a novel concept:
\emph{fractional types}. In the next section, we introduce the problem
of ancilla, motivate its importance, and explain the limitations of
current solutions. In Sec.~\ref{sec:pi}, we review previous work by
James et al.~\cite{rc2011,rc2012,James:2012:IE:2103656.2103667} that
introduces a reversible programming language built using isomorphism
of types. Although it is conceivable that the concept of fractional
types would apply to general-purpose languages (after some
adaptation), its natural technical definition is using symmetries
present in categorical model of type isomorphisms. We present such a
definition in Sec.~\ref{sec:dyn}: that definition captures the main
ideas of fractional types and partially solves a more general and more
expressive instance of the ancilla problem. The approach however
shares the limitation of current solutions in that it relies on
runtime checks for programmer's assertions that values have been
restored to their initial value before de-allocation. To remedy this
limitation, we refine the type system in Sec.~\ref{sec:dep} to keep
track of values in focus. Technically this requires the introduction
of pointed types on top of which are defined both a monad and co-monad
of singleton types. To justify the claim that fractional types solve
the management of ancilla-based scratch space, we present in
Sec.~\ref{sec:space} an operational semantics that exposes memory
management details using a realization of fractional types as
first-class dependently-typed garbage collectors. In Sec.~\ref{cat} we
present a rich collection of examples illustrating the expressive
power of fractional types. We conclude with a discussion of related
work and a summary of our results.

>>>>

\paragraph*{Conservation of Information and Negative Entropy.}  
According to the conventional theory of
communication~\cite{Shannon1948}, a type with $N$ values is viewed as
an abstract system that has $N$ distinguishable states where the
amount of information contained in each state is $\log{N}$. This
entropy is a measure of information which materializes itself in
memory or bandwidth requirements when storing or transmitting elements
of this type. Thus a type with 8 elements needs 3 bits of memory for
storage or 3 bits of bandwidth for communication. The logarithmic map
implies that information contained in a composite state is the sum of
the information contained in its constituents. For example, the type
$A \times B$ where $A$ has two elements and $B$ has three elements can
be thought of a composite system consisting of two independent
unrelated subsystems.  Each state of the composite system therefore
contains $\log{(2*3)} = \log{2} + \log{3}$ bits which is the sum of
the information contained in each subsystem. If we could imagine a
\emph{fractional type} $\frac{1}{A}$, this type would introduce
negative entropy. For example, a type with ``cardinality''
$\frac{1}{8}$ has entropy $\log{\frac{1}{8}} = -3$. It is natural to
interpret this negative entropy just like we interpret ``negative
money,'' as a resource (space or bandwidth) to be repaid (reclaimed)
by some other part of the system. Indeed, we will introduce such
fractional types and use them to represent ``garbage collection
processes'' that reclaim temporary storage. Just like in the case of
``negative money'' it is critical that these fractional types be
linearly used, i.e., they cannot be duplicated or erased: this
property is however automatically enforced in a reversible
computational model.

\paragraph*{Outline.} The paper solves the ancilla problem in
reversible and quantum computation using a novel concept:
\emph{fractional types}. In the next section, we introduce the problem
of ancilla, motivate its importance, and explain the limitations of
current solutions. In Sec.~\ref{sec:pi}, we review previous work by
James et al.~\cite{rc2011,rc2012,James:2012:IE:2103656.2103667} that
introduces a reversible programming language built using isomorphism
of types. Although it is conceivable that the concept of fractional
types would apply to general-purpose languages (after some
adaptation), its natural technical definition is using symmetries
present in categorical model of type isomorphisms. We present such a
definition in Sec.~\ref{sec:dyn}: that definition captures the main
ideas of fractional types and partially solves a more general and more
expressive instance of the ancilla problem. The approach however
shares the limitation of current solutions in that it relies on
runtime checks for programmer's assertions that values have been
restored to their initial value before de-allocation. To remedy this
limitation, we refine the type system in Sec.~\ref{sec:dep} to keep
track of values in focus. Technically this requires the introduction
of pointed types on top of which are defined both a monad and co-monad
of singleton types. To justify the claim that fractional types solve
the management of ancilla-based scratch space, we present in
Sec.~\ref{sec:space} an operational semantics that exposes memory
management details using a realization of fractional types as
first-class dependently-typed garbage collectors. In Sec.~\ref{sec:cat} we
present a rich collection of examples illustrating the expressive
power of fractional types. We conclude with a discussion of related
work and a summary of our results.
